Case study validations of automatic bullet matching

### 2018-11-30 index.Rmd
couple more edits

# to produce blinded version set to 1
blinded: 1
# This gets changed automatically when the title page is compiled in the chunk child-doc-compile
titlepage: 0
# Set to 1 if figures should be included with the text
draft: 0

@@

thanks: 
- The authors wish to thank Jim Hamby, Tylor Klep (Phoenix PD), and Melissa Nally and Kasi Kirksey (FSI Houston) for access to test set bullets. 
- XXX Scanning. 
- Thanks to Jim Hamby for his advice on XXX.

@@

keywords:
- forensic science
- toolmark
- cross-correlation
- random forest
- 3D microscopy
- land engraved areas (LEAs)

abstract: Recent advances in microscopy have made it possible to collect 3D topographic data, enabling virtual comparisons based on the collected 3D data next to traditional comparison microscopy. Automatic matching algorithms have been introduced for various scenarios, such as matching cartridge cases (Tai and Eddy 2018) or matching bullet striae (Hare et al. 2017b,  Chu et al 2013,  De Kinder and Bonfanti 1999). One key aspect of validating automatic matching algorithms is to evaluate the performance of the algorithm on external tests. Here, we are presenting a discussion of the performance of the matching algorithm (Hare et al. 2017b) in three studies. We are considering matching performance based on the Random forest score, cross-correlation, and consecutive matching striae (CMS) at the land-to-land level and, using Sequential Average Maxima scores, also at the bullet-to bullet level. Cross-correlation and Random Forest scores both result in perfect discrimination of same-source and different-source bullets. At the land-to-land level, discrimination (based on area under the curve, AUC) is excellent ($ge$ 0.90).

@@

More explicit characterization of bullet surfaces has been discussed since at least 1958 by \citeauthor{davis1968introduction} \cite{davis1968introduction}, but at the time technology was not sufficiently advanced to make an analysis based on 3D measurements or surface traces a viable option. \citet{ComputerIdentificationBullets1978} demonstrated use of a scanning electron microscope to quantitatively examine and compare bullet striae. More recently, approaches using 3D measurement data were explicitly described in 1999 (\citeauthor{dekinderAutomatedComparisonsBullet1999}), and have been further developed in \citep{bachrachDevelopment3DbasedAutomated2002,xieAutomatedBulletidentificationSystem2009, chuPilotStudyAutomated2010}. These approaches utilize 3D surface measurements directly to characterize the topology of land engraved areas (LEAs), rather than using visual or image comparison techniques. In many cases, these approaches also provide some level of automation of the comparison process, with the goal of reducing human biases by augmenting the visual information with 3D measurements. Utilizing the 3D measurements allows for examination of both peaks and valleys in LEAs, it also allows us to take the depth of striae into account, something that is difficult, if not impossible, from visual inspection.

Commonly, approaches derived from 3D surface measurements use some features which are similar to those visually assessed by FTEs \citep{luAutomatedBulletIdentification2014}. Class characteristics, which are shared by a group of firearms with the same rifling design, manufacturer, and tooling process, are typically evaluated first, as a mismatch on class charcteristics is sufficient for an exclusion. Automated approaches to estimation of width of the land engraved area and twist angle were described in \citet{chuPilotStudyAutomated2010}. Individual characteristics, which are not shared by all members of a class, can also be automatically assessed from 3D surface measurements. One of the most common features used to assess the strength of a match based on individual characteristics is the cross-correlation function, which is utilized in several studies \citep{maNISTBulletSignature2004, vorburgerApplicationsCrosscorrelationFunctions2011, chuPilotStudyAutomated2010}. Additional features proposed for automatically assessing match strength also include signature distance \citep{maNISTBulletSignature2004}, striae depth and width \citep{ComputerIdentificationBullets1978}, and consecutive matching striae (CMS) \citep{chuAutomaticIdentificationBullet2013}.

@@

In the matching algorithm proposed in \citet{aoas}, several features are combined using a random forest \citep{breiman} to produce a similarity score based on 3D topographic scans of land engraved areas (LEAs). \svp{In order to generate these features, some pre-processing is necessary in order to transform the 3D surface measurements into `signatures' which can be compared.}

@@

Once signatures have been extracted from stable regions with expressed striae, they must be aligned in order to assess their match strength. Maximized cross-correlation is used to pair-wise align signatures \citep{aoas,vorburgerApplicationsCrosscorrelationFunctions2011}.

The cross-correlation (CC) function between two signatures $S_1(x)$ and $S_2(x)$ is defined as 
\[
CC(S_1, S_2, z) = \text{cor}\left(S_1(x), S_2(x + z)\right)
\] 
where $x$ and $x+z$ are integer values appropriately defined within the domains of $S_1$ and $S_2$, $z$ is the lag between the two sequences, \svp{and }$\text{cor}(.,.)$ is the Pearson correlation coefficient.
When signatures contain missing values, pairwise complete observations are used to calculate the cross-correlation. The lag $z$ used to achieve maximum $CC$ is used to determine the best alignment.

@@

Using the lag determined for aligning two signatures, other quantitative features describing the similarity of the two signatures can be extracted. Numerical features such as cross-correlation and Euclidean distance
can be computed from the aligned signatures alone. Other features include characteristics of the signature that firearms and toolmark examiners base their visual assessment on, such as striae depth, number of matching striae, and the number of consecutively matching striae (CMS), one of the measures firearms examiners use to describe the strength of a match quantitatively.

In order to evaluate CMS, we must first identify peaks and valleys in each of the signatures, then determine whether these peaks and valleys overlap sufficiently. Features which firearms examiners assess visually depend on the identification of these extrema. If there are at least six consecutive matching striae, the bullets are considered to be a good match in practice \citep{biasotti_firearms_1997,nichols_scientific_2006}. CMS in this setting are counted as the number of consecutively matching peaks in signatures of two aligned lands. Peaks -- rather than peaks and valleys -- are used for this determination in part because it is difficult to visually assess valleys except as relative to peaks. Note that with automatic alignment based on the maximum cross-correlation rather than matching striation marks, it is possible for two lands to have a consecutive matching striae of zero.

<!-- I really think this belongs in this section: it's discussing cross-correlation as a scoring statistic, rather than as a convenient way to align signatures so that scoring stats can be calculated. -->

\svp{Cross correlation can also be used to assess match strength between two aligned signatures. Cross correlation varies between -1 and 1, but as we are taking the maximal value, in practice the cross-correlations for aligned signatures are generally positive. Cross-correlation values which are close to 1 indicate similarity between signatures and with it may indicate that the two signatures come from the same source. Cross-correlation values which are close to 0 indicate signatures which may originate from different sources.}

The random forest presented in \citet{aoas} is based on a combination of multiple characteristics, including cross-correlation, number of matching striae, and number of consecutively matching striae. The output of the random forest is a score between 0 and 1 representing the algorithm's assessment of the strength of the match, where scores close to 1 indicate it to be highly likely that the aligned signatures come from the same source, while scores close to 0 indicate that the aligned signatures are highly likely to originate from different sources.

@@

\svp{The features described above are each individually designed to separate same-source and different-source signatures: for instance, same-source signatures would have high cross-correlation values and different-source signatures would have low cross-correlation values. An automated algorithm selects a cutoff value, introducing a barrier between identification and exclusion.} 
<!-- rephrased? trying to get rid of "supposed" and "same sourceness" :-p

\hh{Any of the above described features are supposed to separate between signatures from same source and different source, e.g.\ high cross-correlation values are indicative of same sourceness of two signatures, whereas  low values are more indicate of different sources. 
An automated algorithm determines a cutoff-value along the range of possible values that introduces a barrier between identification and exclusion.
}
-->

<!-- Firearms examiners typically set the threshold for CMS at 6; this is a threshold which is conservative, as it is much more likely to exclude same-source comparisons than it is to identify different-source comparisons.  -->

@@

\begin{definition}{Sequence Average and its Maximum}
Let $A$ be a square real-valued matrix of dimensions $p \times p$. For the purpose of this paper, $A$ consists of scores describing the similarity between two sets of land engraved areas.
The $k$th \emph{sequence average} $SA(A, k)$ for $k = 0, ... p-1$ is defined  as 
$$SA(A, k) = \frac{1}{p} \sum_{i=1}^{p} a_{i,i \oplus k}, \text{ where } i\oplus k := \left((i + k - 1)\mod p\right) + 1.$$
The \emph{Sequence Average Maximum} \citep[SAM, ][]{sam} of square matrix $A$ of scores is defined as
$$SAM (A) = \max_{k = 1}^{p} SA(A, k).$$
\end{definition}
\svp{XXX Rearranged this slightly XXX}
Looking back at \autoref{fig:b2b}, we see that for the two bullets from the same barrel, the sequence average for 
$k=2$ is higher than the other sequence averages, and also higher than the sequence averages for the other pair of bullets shown on the right of the figure. SAM scores allow us to define a single quantity for each pair of bullets that describes the similarity between these two bullets.

@@

In this paper, we validate the algorithm described in \citet{aoas} on three external test sets. We assess the results on each set and evaluate whether the minimal requirements described above are fulfilled for bullet-to-bullet SAM scores. We compare the random forest algorithm's performance to the performance of other suggested measures for quantitative assessment of match strength, such as cross-correlation and consecutive matching striae.

@@

# Evaluating the Random Forest Algorithm
The random forest SAM scores correctly separate known matches from known non-matches, using SAM scores to aggregate land-to-land scores into a bullet-to-bullet comparison. With no errors at the bullet-to-bullet level, we can now evaluate the scores in light of (R1) and (R2). In this section, we will compare the random forest scores to cross-correlation scores; cross-correlation is one of the components of the random forest, but has also been used to quantify the strength of a match in its own right \citep{chuPilotStudyAutomated2010}. In addition, both scoring methods use the same scale (0 to 1) and are continuous along that interval; other evaluation methods, such as consecutive matching striae are discrete and more difficult to directly compare to continuous measurements.

## Comparison of SAM scores
\autoref{fig:compare} shows density curves for the scores bullet-to-bullet matches of each of the studies, comparing the SAM scores based on cross-correlation (top) and the random forest score (bottom). Color indicates ground truth. Small vertical lines below the $x$-axis indicate observed scores in a **rug-plot**. The plot shows that for all three case studies and both continuous measures, i.e.\ SAM scores based on cross-correlation as well as SAM scores based on the random forest scores, all similarity scores are higher for pairs of bullets from the same source, i.e.\ bullets fired through the same barrel, than pairs of bullets from different sources (fired through different barrels).

```{r compare, fig.width = 8, fig.height = 5, fig.cap="Density curves of similarity scores from cross-correlation (top) and RF scores (bottom). Different colors indicate same source versus different source. Ideally all scores of different source pairs should be much lower than scores for same source pairs."}
nests <- rbind(
  h44nest %>%
    select(sam_rf, sam_ccf, samesource) %>%
    mutate(study = "Hamby 44", group1 = NA),
  pdnest %>%
    select(sam_rf, sam_ccf, samesource) %>%
    mutate(study = "Phoenix PD", group1 = NA),
  hounest %>%
    select(sam_rf, sam_ccf,  samesource, group1) %>%
    mutate(study = "Houston FSI")
)
nests$study <- factor(nests$study, levels = c("Hamby 44", "Phoenix PD", "Houston FSI"))
nestslong <- nests %>% gather(key = "measure", value = "value", sam_rf, sam_ccf)
nestslong$measure <- factor(nestslong$measure)
levels(nestslong$measure) <- c("Cross-correlation",  "RF score")

@@

According to \autoref{fig:compare}, neither cross-correlation nor random forest score fulfill both of the minimal requirements laid out in the introduction. Both measures fulfill (R1), i.e.\ all  scores  of  different-source pairs are lower than scores of same-source pairs for SAM scores based on cross-correlation and RF scores.
However, SAM scores based on the multivariate random forest score show better separation between same-source and different-source scores than SAM scores based on cross-correlation. This can be seen from the larger horizontal distance between the modes (peaks) of the density curves of RF based SAM scores compared to the cross-correlation SAM scores.

@@

Examining land-to-land scores provides more details about the matching algorithm's performance. \autoref{fig:compare-land-to-land} shows density curves for each study. There is some overlap between known matching and known non-matching land-to-land scores; in many cases the random forest scores for these overlapping values are more extreme than the corresponding cross-correlation scores. This suggests that the matching algorithm is sensitive to the presence of LEAs which have low land-to-land match scores, such as the matches of bullets 'I' and 'F' in the Hamby 44 study. Comparing this to the bullet-to-bullet score density plot shown in \autoref{fig:compare}, we see that there is much greater separation between the same-source and different-source densities for each study in the bullet-to-bullet comparisons than in the land-to-land comparisons, that is, one or two poorly matching lands does not prevent the bullet from showing a matching score, though if there are weak matches on several lands, such as the marked Hamby 44 comparisons in \autoref{fig:compare}, the overall score may be affected.

@@

```{r compare-land-to-land, fig.width = 8, fig.height = 4, fig.cap="Density curves of land-to-land similarity scores from cross-correlation (top) and RF scores (bottom). Different colors indicate same source versus different source for each land. RF scores for different source comparisons are generally well below 0.5. Some same-source comparisons are also below 0.5 for RF scores, indicating potential problems with at least one of the lands involved in the comparison."}
all_lands %>% 
  gather(key = "measure", value = "value", ccf, rfscore) %>%
  mutate(measure = factor(measure, levels = c("ccf", "rfscore"),
                          labels = c("Cross-correlation", "RF score")),
         study = paste(study, group1) %>% str_trim() %>% factor(levels = c("Hamby 44", "Phoenix PD", "Houston FSI G1", "Houston FSI G2", "Houston FSI G3"))) %>%
  ggplot(aes(
    x = value, fill = sameland,
    colour = sameland
  )) +
  geom_density(alpha = 0.6) +
#  geom_rug(alpha = 0.5/12, show.legend = F) +
  facet_grid(measure ~ study) +
  theme_bw() +
  scale_fill_manual("Same source", values = c("darkgrey", "darkorange")) +
  scale_colour_manual("Same source", values = c("darkgrey", "darkorange")) +
  scale_x_continuous("Land-to-Land scores", limits = c(0, 1), labels = c(0, 0.25, 0.5, 0.75, 1)) +
  theme(legend.position = "bottom") +
  geom_hline(yintercept = 0, show.legend = F)
```

@@

```{r roc-auc, echo = FALSE, out.width = c("0.51\\textwidth", ".4725\\textwidth"), out.height = c("0.63\\textwidth", "0.63\\textwidth"), fig.width = c(5, 6), fig.height = c(7.5, 7.5), fig.subcap=c("ROC curves", "Area under the curve (AUC)"), fig.cap="ROC curves and AUC for each (set) of the studies for the random forest score (RF), cross-correlation (CC) and consecutively matching striae (CMS).   As seen in the ROC curves, the algorithm performs the least well on set 2 of the Houston FSI study. Based on AUC, the overall performance on all sets is very good to excellent for cross-correlation and the random forest score. At the land level there is no significant difference in prediction power between RF score and cross-correlation.", cache=FALSE}
  rocpic
  aucpic
```

In \autoref{fig:roc-auc}a, the Houston FSI G1 and G3 curves and the Phoenix PD curve show excellent performance; Houston FSI G2 and Hamby 44 show still a very  good performance. Area under the curve (AUC) values, which summarize ROC curves, are shown in \autoref{fig:roc-auc}b. AUC values are useful for differentiating between poor, good, and excellent model performance, but are not particularly useful when determining which of several models with approximately the same level of performance should be used \citep{marzban_roc_2004}. Hamby 44 and Phoenix PD are the only studies with a somewhat noticable difference between the AUC score for RF compared to cross-correlation, but even those differences are not statistically significant.

\autoref{fig:eer} shows an overview of Equal Error and their thresholds based on the ROC curves of \autoref{fig:roc-auc}. Equal Errors are errors when sensitivity and specificity of a test are equal, i.e.\ we see the same percentage of missed and wrong identifications for a land-to-land comparison. Equal errors based on CMS of two lands are significantly higher than equal errors based on cross-correlation or the RF score. At best, RF score and cross-correlation have an equal error of around 5\% for land-to-land comparisons. 
Interestingly, for consecutively matching striae, equal errors are reached for just two matching striae, which from a practical point of view is not sustainable. A  CMS of  6 or above is an established cutoff used in identifications; when applied to the land-to-land comparisons, it is much more likely to result in a missed identification than a wrong identification. A threshold of 6 leads to 21 wrong identifications out of 49,286 different-source comparisons (0.04\% errors, or about 4 errors for each 10,000 evaluations); when applied to same-source comparisons, 517 of 715 same-source lands were incorrectly classified as exclusions, for an error rate of 80.6\%.

@@

```{r eer, fig.width=7, fig.height = 4, fig.cap="Equal Error (in Percent)  and corresponding thresholds for all measures (RF score, Cross-correlation and CMS) under all five (sub-)studies. CMS has an equal error threshold of 2 for all studies. CMS is signifcantly worse than both Cross-correlation and the RF score. At best, Cross-correlation and RF score have equal errors for land-to-land comparison of around 5\\%.", out.width = '.8\\textwidth'}
aucs %>% 
  mutate(
    eer = 100*eer
  ) %>%
  gather(eer_thresh, value, eer, threshold) %>%
  filter(!(measure=="CMS" & eer_thresh=="threshold")) %>%
  mutate(
    eer_thresh = factor(eer_thresh, labels=c("Equal Error (in Percent)", "Threshold for Equal Error"))
  ) %>%
  ggplot(aes(y = studyg, x = value, colour=measure, shape = measure)) + 
  geom_point(size = 3.5) +
  theme_bw() +
  xlab("Equal Error (in Percent, left) and corresponding thresholds (right)") +
  ylab("") +
  scale_color_brewer(palette = "Set2") +
  scale_shape_discrete(solid = F) + 
  theme(legend.position = "bottom") +
  facet_grid(.~eer_thresh, scales="free") +
  xlim(c(0, NA))
```

>>>>

# to produce blinded version set to 1
blinded: 0
# This gets changed automatically when the title page is compiled in the chunk child-doc-compile
titlepage: 1
# Set to 1 if figures should be included with the text
draft: 1

@@

thanks: 
- The authors wish to thank Alan Zheng (NIST), Tylor Klep (Phoenix PD), and Melissa Nally and Kasi Kirksey (FSI Houston) for access to test set bullets. 
- We would like to thank the efforts of the Roy J Carver High Resolution lab in scanning the bullet sets and providing the scans to us. 
- Thanks to Prof James Hamby for his advice on XXX.

@@

keywords:
- forensic science
- toolmark
- cross correlation
- random forest
- 3D microscopy
- land engraved areas (LEAs)

abstract: Recent advances in microscopy have made it possible to collect 3D topographic data, enabling virtual comparisons based on the collected 3D data next to traditional comparison microscopy. Automatic matching algorithms have been introduced for various scenarios, such as matching cartridge cases (Tai and Eddy 2018) or matching bullet striae (Hare et al. 2017b,  Chu et al 2013,  De Kinder and Bonfanti 1999). One key aspect of validating automatic matching algorithms is to evaluate the performance of the algorithm on external tests. Here, we are presenting a discussion of the performance of the matching algorithm (Hare et al. 2017b) in three studies. We are considering matching performance based on the Random forest score, cross corrleation, and consecutive matching striae (CMS) at the land-to-land level and, using Sequential Average Maxima scores, also at the bullet-to bullet level. Cross corrleation and Random Forest scores both result in perfect discrimination of same-source and different-source bullets. At the land-to-land level, discrimination (based on area under the curve, AUC) is excellent ($ge$ 0.90).

@@

More explicit characterization of bullet surfaces has been discussed since at least 1958 by \citeauthor{davis1968introduction} \cite{davis1968introduction}, but at the time technology was not sufficiently advanced to make an analysis based on 3D measurements or surface traces a viable option. \citet{ComputerIdentificationBullets1978} demonstrated use of a scanning electron microscope to quantitatively examine and compare bullet striae. More recently, approaches using 3D measurement data were explicitly described in 1999 by \citeauthor{dekinderAutomatedComparisonsBullet1999} \cite{dekinderAutomatedComparisonsBullet1999}, and have been further developed in \citep{bachrachDevelopment3DbasedAutomated2002,xieAutomatedBulletidentificationSystem2009, chuPilotStudyAutomated2010}. These approaches utilize 3D surface measurements directly to characterize the topology of land engraved areas (LEAs), rather than using visual or image comparison techniques. In many cases, these approaches also provide some level of automation of the comparison process, with the goal of reducing human biases by augmenting the visual information with 3D measurements. Utilizing the 3D measurements allows for examination of both peaks and valleys in LEAs, it also allows us to take the depth of striae into account, something that is difficult, if not impossible, from visual inspection.

Commonly, approaches derived from 3D surface measurements use some features which are similar to those visually assessed by FTEs \citep{luAutomatedBulletIdentification2014}. Class characteristics, which are shared by a group of firearms with the same rifling design, manufacturer, and tooling process, are typically evaluated first, as a mismatch on class charcteristics is sufficient for an exclusion. Automated approaches to estimation of width of the land engraved area and twist angle were described in \citet{chuPilotStudyAutomated2010}. Individual characteristics, which are not shared by all members of a class, can also be automatically assessed from 3D surface measurements. One of the most common features used to assess the strength of a match based on individual characteristics is the cross corrleation function, which is utilized in several studies \citep{maNISTBulletSignature2004, vorburgerApplicationsCrosscorrelationFunctions2011, chuPilotStudyAutomated2010}. Additional features proposed for automatically assessing match strength also include signature distance \citep{maNISTBulletSignature2004}, striae depth and width \citep{ComputerIdentificationBullets1978}, and consecutive matching striae (CMS) \citep{chuAutomaticIdentificationBullet2013}.

@@

In the matching algorithm proposed in \citet{aoas}, several features are combined using a random forest \citep{breiman} to produce a similarity score based on 3D topographic scans of land engraved areas (LEAs). In order to generate these features, some pre-processing is necessary in order to transform the 3D surface measurements into `signatures' which can be compared.

@@

Once signatures have been extracted from stable regions with expressed striae, they must be aligned in order to assess their match strength. Maximized cross corrleation is used to pair-wise align signatures \citep{aoas,vorburgerApplicationsCrosscorrelationFunctions2011}.

The cross corrleation (CC) function between two signatures $S_1(x)$ and $S_2(x)$ is defined as 
\[
CC(S_1, S_2, z) = \text{cor}\left(S_1(x), S_2(x + z)\right)
\] 
where $x$ and $x+z$ are integer values appropriately defined within the domains of $S_1$ and $S_2$, $z$ is the lag between the two sequences, and $\text{cor}(.,.)$ is the Pearson correlation coefficient.
When signatures contain missing values, pairwise complete observations are used to calculate the cross corrleation. The lag $z$ used to achieve maximum $CC$ is used to determine the best alignment.

@@

Using the lag determined for aligning two signatures, other quantitative features describing the similarity of the two signatures can be extracted. Numerical features such as cross corrleation and Euclidean distance
can be computed from the aligned signatures alone. Other features include characteristics of the signature that firearms and toolmark examiners base their visual assessment on, such as striae depth, number of matching striae, and the number of consecutively matching striae (CMS), one of the measures firearms examiners use to describe the strength of a match quantitatively.

In order to evaluate CMS, we must first identify peaks and valleys in each of the signatures, then determine whether these peaks and valleys overlap sufficiently. Features which firearms examiners assess visually depend on the identification of these extrema. If there are at least six consecutive matching striae, the bullets are considered to be a good match in practice \citep{biasotti_firearms_1997,nichols_scientific_2006}. CMS in this setting are counted as the number of consecutively matching peaks in signatures of two aligned lands. Peaks -- rather than peaks and valleys -- are used for this determination in part because it is difficult to visually assess valleys except as relative to peaks. Note that with automatic alignment based on the maximum cross corrleation rather than matching striation marks, it is possible for two lands to have a consecutive matching striae of zero.

<!-- I really think this belongs in this section: it's discussing cross corrleation as a scoring statistic, rather than as a convenient way to align signatures so that scoring stats can be calculated. -->

Cross correlation can also be used to assess match strength between two aligned signatures. Cross correlation varies between -1 and 1, but as the alignment is based on the maximized cross correlation value, in practice the cross correlations for aligned signatures are generally positive. Cross correlation values which are close to 1 indicate similarity between signatures and with it may indicate that the two signatures come from the same source. Low cross corrleation values indicate signatures which may originate from different sources.

The random forest presented in \citet{aoas} is based on a combination of multiple characteristics, including cross corrleation, number of matching striae, and number of consecutively matching striae. The output of the random forest is a score between 0 and 1 representing the algorithm's assessment of the strength of the match, where scores close to 1 indicate it to be highly likely that the aligned signatures come from the same source, while scores close to 0 indicate that the aligned signatures are highly likely to originate from different sources.

@@

The features described above are each individually designed to separate same-source and different-source signatures: for instance, same-source signatures would have high cross corrleation values and different-source signatures would have low cross corrleation values. An automated algorithm selects a cutoff value, introducing a barrier between identification and exclusion.

@@

\begin{definition}{Sequence Average and its Maximum}
Let $A$ be a square real-valued matrix of dimensions $p \times p$. For the purpose of this paper, $A$ consists of scores describing the similarity between two sets of land engraved areas.
The $k$th \emph{sequence average} $SA(A, k)$ for $k = 0, ... p-1$ is defined  as 
$$SA(A, k) = \frac{1}{p} \sum_{i=1}^{p} a_{i,i \oplus k}, \text{ where } i\oplus k := \left((i + k - 1)\mod p\right) + 1.$$
The \emph{Sequence Average Maximum} \citep[SAM, ][]{sam} of square matrix $A$ of scores is defined as
$$SAM (A) = \max_{k = 1}^{p} SA(A, k).$$
\end{definition}
Looking back at \autoref{fig:b2b}, we see that for the two bullets from the same barrel, the sequence average for 
$k=2$ is higher than the other sequence averages, and also higher than the sequence averages for the other pair of bullets shown on the right of the figure. SAM scores allow us to define a single quantity for each pair of bullets that describes the similarity between these two bullets.

@@

In this paper, we validate the algorithm described in \citet{aoas} on three external test sets. We assess the results on each set and evaluate whether the minimal requirements described above are fulfilled for bullet-to-bullet SAM scores. We compare the random forest algorithm's performance to the performance of other suggested measures for quantitative assessment of match strength, such as cross corrleation and consecutive matching striae.

@@

# Evaluating the Random Forest Algorithm
The random forest SAM scores correctly separate known matches from known non-matches, using SAM scores to aggregate land-to-land scores into a bullet-to-bullet comparison. With no errors at the bullet-to-bullet level, we can now evaluate the scores in light of (R1) and (R2). In this section, we will compare the random forest scores to cross corrleation scores; cross corrleation is one of the components of the random forest, but has also been used to quantify the strength of a match in its own right \citep{chuPilotStudyAutomated2010}. In addition, both scoring methods use the same scale (0 to 1) and are continuous along that interval; other evaluation methods, such as consecutive matching striae are discrete and more difficult to directly compare to continuous measurements.

## Comparison of SAM scores
\autoref{fig:compare} shows density curves for the scores bullet-to-bullet matches of each of the studies, comparing the SAM scores based on cross corrleation (top) and the random forest score (bottom). Color indicates ground truth. Small vertical lines below the $x$-axis indicate observed scores in a **rug-plot**. The plot shows that for all three case studies and both continuous measures, i.e.\ SAM scores based on cross corrleation as well as SAM scores based on the random forest scores, all similarity scores are higher for pairs of bullets from the same source, i.e.\ bullets fired through the same barrel, than pairs of bullets from different sources (fired through different barrels).

```{r compare, fig.width = 8, fig.height = 5, fig.cap="Density curves of similarity scores from cross corrleation (top) and RF scores (bottom). Different colors indicate same source versus different source. Ideally all scores of different source pairs should be much lower than scores for same source pairs."}
nests <- rbind(
  h44nest %>%
    select(sam_rf, sam_ccf, samesource) %>%
    mutate(study = "Hamby 44", group1 = NA),
  pdnest %>%
    select(sam_rf, sam_ccf, samesource) %>%
    mutate(study = "Phoenix PD", group1 = NA),
  hounest %>%
    select(sam_rf, sam_ccf,  samesource, group1) %>%
    mutate(study = "Houston FSI")
)
nests$study <- factor(nests$study, levels = c("Hamby 44", "Phoenix PD", "Houston FSI"))
nestslong <- nests %>% gather(key = "measure", value = "value", sam_rf, sam_ccf)
nestslong$measure <- factor(nestslong$measure)
levels(nestslong$measure) <- c("Cross corrleation",  "RF score")

@@

According to \autoref{fig:compare}, neither cross corrleation nor random forest score fulfill both of the minimal requirements laid out in the introduction. Both measures fulfill (R1), i.e.\ all  scores  of  different-source pairs are lower than scores of same-source pairs for SAM scores based on cross corrleation and RF scores.
However, SAM scores based on the multivariate random forest score show better separation between same-source and different-source scores than SAM scores based on cross corrleation. This can be seen from the larger horizontal distance between the modes (peaks) of the density curves of RF based SAM scores compared to the cross corrleation SAM scores.

@@

Examining land-to-land scores provides more details about the matching algorithm's performance. \autoref{fig:compare-land-to-land} shows density curves for each study. There is some overlap between known matching and known non-matching land-to-land scores; in many cases the random forest scores for these overlapping values are more extreme than the corresponding cross corrleation scores. This suggests that the matching algorithm is sensitive to the presence of LEAs which have low land-to-land match scores, such as the matches of bullets 'I' and 'F' in the Hamby 44 study. Comparing this to the bullet-to-bullet score density plot shown in \autoref{fig:compare}, we see that there is much greater separation between the same-source and different-source densities for each study in the bullet-to-bullet comparisons than in the land-to-land comparisons, that is, one or two poorly matching lands does not prevent the bullet from showing a matching score, though if there are weak matches on several lands, such as the marked Hamby 44 comparisons in \autoref{fig:compare}, the overall score may be affected.

@@

```{r compare-land-to-land, fig.width = 8, fig.height = 4, fig.cap="Density curves of land-to-land similarity scores from cross corrleation (top) and RF scores (bottom). Different colors indicate same source versus different source for each land. RF scores for different source comparisons are generally well below 0.5. Some same-source comparisons are also below 0.5 for RF scores, indicating potential problems with at least one of the lands involved in the comparison."}
all_lands %>% 
  gather(key = "measure", value = "value", ccf, rfscore) %>%
  mutate(measure = factor(measure, levels = c("ccf", "rfscore"),
                          labels = c("Cross corrleation", "RF score")),
         study = paste(study, group1) %>% str_trim() %>% factor(levels = c("Hamby 44", "Phoenix PD", "Houston FSI G1", "Houston FSI G2", "Houston FSI G3"))) %>%
  ggplot(aes(
    x = value, fill = sameland,
    colour = sameland
  )) +
  geom_density(alpha = 0.6) +
#  geom_rug(alpha = 0.5/12, show.legend = F) +
  facet_grid(measure ~ study) +
  theme_bw() +
  scale_fill_manual("Same source", values = c("darkgrey", "darkorange")) +
  scale_colour_manual("Same source", values = c("darkgrey", "darkorange")) +
  scale_x_continuous("Land-to-Land scores", limits = c(0, 1), labels = c(0, 0.25, 0.5, 0.75, 1)) +
  theme(legend.position = "bottom") +
  geom_hline(yintercept = 0, show.legend = F)
```

@@

```{r roc-auc, echo = FALSE, out.width = c("0.51\\textwidth", ".4725\\textwidth"), out.height = c("0.63\\textwidth", "0.63\\textwidth"), fig.width = c(5, 6), fig.height = c(7.5, 7.5), fig.subcap=c("ROC curves", "Area under the curve (AUC)"), fig.cap="ROC curves and AUC for each (set) of the studies for the random forest score (RF), cross corrleation (CC) and consecutively matching striae (CMS).   As seen in the ROC curves, the algorithm performs the least well on set 2 of the Houston FSI study. Based on AUC, the overall performance on all sets is very good to excellent for cross corrleation and the random forest score. At the land level there is no significant difference in prediction power between RF score and cross corrleation.", cache=FALSE}
  rocpic
  aucpic
```

In \autoref{fig:roc-auc}a, the Houston FSI G1 and G3 curves and the Phoenix PD curve show excellent performance; Houston FSI G2 and Hamby 44 show still a very  good performance. Area under the curve (AUC) values, which summarize ROC curves, are shown in \autoref{fig:roc-auc}b. AUC values are useful for differentiating between poor, good, and excellent model performance, but are not particularly useful when determining which of several models with approximately the same level of performance should be used \citep{marzban_roc_2004}. Hamby 44 and Phoenix PD are the only studies with a somewhat noticable difference between the AUC score for RF compared to cross corrleation, but even those differences are not statistically significant.

\autoref{fig:eer} shows an overview of Equal Error and their thresholds based on the ROC curves of \autoref{fig:roc-auc}. Equal Errors are errors when sensitivity and specificity of a test are equal, i.e.\ we see the same percentage of missed and wrong identifications for a land-to-land comparison. Equal errors based on CMS of two lands are significantly higher than equal errors based on cross corrleation or the RF score. At best, RF score and cross corrleation have an equal error of around 5\% for land-to-land comparisons. 
Interestingly, for consecutively matching striae, equal errors are reached for just two matching striae, which from a practical point of view is not sustainable. A  CMS of  6 or above is an established cutoff used in identifications; when applied to the land-to-land comparisons, it is much more likely to result in a missed identification than a wrong identification. A threshold of 6 leads to 21 wrong identifications out of 49,286 different-source comparisons (0.04\% errors, or about 4 errors for each 10,000 evaluations); when applied to same-source comparisons, 517 of 715 same-source lands were incorrectly classified as exclusions, for an error rate of 80.6\%.

@@

```{r eer, fig.width=7, fig.height = 4, fig.cap="Equal Error (in Percent)  and corresponding thresholds for all measures (RF score, Cross corrleation and CMS) under all five (sub-)studies. CMS has an equal error threshold of 2 for all studies. CMS is signifcantly worse than both Cross corrleation and the RF score. At best, Cross corrleation and RF score have equal errors for land-to-land comparison of around 5\\%.", out.width = '.8\\textwidth'}
aucs %>% 
  mutate(
    eer = 100*eer
  ) %>%
  gather(eer_thresh, value, eer, threshold) %>%
  filter(!(measure=="CMS" & eer_thresh=="threshold")) %>%
  mutate(
    eer_thresh = factor(eer_thresh, labels=c("Equal Error (in Percent)", "Threshold for Equal Error"))
  ) %>%
  ggplot(aes(y = studyg, x = value, colour=measure, shape = measure)) + 
  geom_point(size = 3.5) +
  theme_bw() +
  xlab("Equal Error (in Percent, left) and corresponding thresholds (right)") +
  ylab("") +
  scale_color_brewer(palette = "Set2") +
  scale_shape_discrete(solid = F) + 
  theme(legend.position = "bottom") +
  facet_grid(.~eer_thresh, scales="free") +
  xlim(c(0, NA))
```
